---
title: "Week 3 R Assignment - Transforming and recoding data"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---
Directions: Complete each step of the assignment below.

Preparation: Run the following code block to install the R packages needed for this assignment and load them into memory. (This may take a few minutes if you haven't installed the packages before. If you get any errors or warnings, try running the block another time.)
```{r}
# List the packages needed
packages <- c("summarytools","ggplot2","Hmisc","pastecs","car")

# Check if each package is installed already. If not, install the package.
for (i in packages){
  if(! i %in% installed.packages()){
    install.packages(i,dependencies = TRUE)
  }
  # Show each package that is checked.
  print(i)
  
  # Load each package into memory so we can use it.
  library(i,character.only=TRUE)
}
```


1) Run the following code block below to open the Week1.csv file.
```{r}
data <- read.csv(file.choose())
```

2) Some csv data sets may have additional characters at the beginning of the csv file which can show up in the first variable name when the data set is imported into R. (This can happen sometimes when the data are exported from Excel into a csv file.) If the first variable name has strange characters at the beginning, run the code block below to rename the first variable.
```{r}
colnames(data)[1] <- "nominal"
names(data)
```

3) Probably one of themost important transformations you should know how to perform is transforming values in a variable to z-scores. I never transform variables in place. I always transform variables into a new variable. That way if you make a mistake you can delete the new variable and try again. If you recode your variable in place and make a mistake, you will lose your original data and need to reload your dataset. Run the following code block and answer the following questions. 
a) What is the formula to calculate a z-score?
b) What is the mean of the new z-score variable?
c) What is the standard deviation of the new z-score variable?
d) What does it mean if a z-score is postive? (What does it tell you about the original raw score?)
e) What does it mean if a z-score is negative? (What does it tell you about the original raw score?)
```{r}
# transform the variable interval to z-scores
data$interval_zscore <- (data$interval - mean(data$interval))/sd(data$interval)
round(mean(data$interval_zscore),2)
sd(data$interval_zscore)

```

4) There is a function in R to calcuate a z-score. Run the following code to calculate the z-scores for interval in another new variable. Compare the values of the two variables.
a) What do you noice about the two scores?
```{r}
data$interval_zscore2 <- scale(data$interval)
```

5) Insert code into the code block below to transform the variable ratio into z-scores.
```{r}
# transform the variable ratio to z-scores

```

6) Create a scatterplot for the variables interval and ratio. Create another scatterplot for the z-scores of each of those variables. Answer the question below.
a) What do you notice about the scatterplots? 
```{r}
# create scatterplots

```

7) We won't discuss more about data transformations now, but you should understand that just like you can enter a formula to create a z-score for all the values in a variable, you can use other formulas. To practice, calculate the square root for each value in the ratio variable and save the results in a new variabl called ratio_sqrt. Some other common transformations you may see used are log or inverse transformations and there are still others.
 
```{r}
# create a new variable ratio_sqrt

```

8) Another common procedure you may need to know when working with study data is reverse scoring questions on a questionnaire. One way to do this is to specify each value in a variable and identify the new value desired in the new variable.
```{r}
data$ordinal_recode1[data$ordinal=="first"] <- 4
data$ordinal_recode1[data$ordinal=="second"] <- 3
data$ordinal_recode1[data$ordinal=="third"] <- 2
data$ordinal_recode1[data$ordinal=="fourth"] <- 1
data$ordinal_recode1[is.na(data$ordinal)] <- NA
str(data$ordinal_recode1)

# don't forget to make sure the new variable is ordinal like the original
data$ordinal_recode1 <- factor(data$ordinal_recode1,levels=c("1","2","3","4"),ordered=TRUE)
str(data$ordinal_recode1)

```

9) Another way you can do the same thing is using the recode() function in the car package. (Companion to Applied Regression)
```{r}
library(car)
data$ordinal_recode2 <- recode(data$ordinal,"'fourth'=1;'third'=2;'second'=3;'first'=4;NA=NA")
str(data$ordinal_recode2)

# Don't forget to make sure the levels are in the correct order for ordinal variables
data$ordinal_recode2 <- factor(data$ordinal_recode2,levels=c("1","2","3","4"),ordered=TRUE)
str(data$ordinal_recode2)
```

10) Run the code block below to calculate the 30th, 60th, and 90th percentile for the ratio variable.
a) 30th - 
b) 60th - 
c) 90th - 
d) What does the percentile mean?
```{r}
# calculate the 30th, 60th, and 90th percentile for the ratio variable
print("percentiles using actual data")
quantile(data$ratio,c(.30,.60,.90))

# qnorm will give you the corresponding z-scores for a given percent rank
print("z scores")
qnorm(c(.30,.60,.90))

# sometimes you don't have the actual data, but you can calculate percentiles using the normal distribution for a given mean and sd
# qnorm will calculate a percentile for a percentile rank if supplied with a mean and sd
print("percentiles using normal distribution given mean and sd")
qnorm(c(.30,.60,.90),mean(data$ratio),sd(data$ratio))

```

11) Calculate the percentile rank for each score in the ratio variable using the z-score.
a) What is the percentile rank for the lowest score?
b) What is the percentile rank forthe highest score?
c) What does the percentile rank mean?
```{r}
data$ratio_rank <- rank(data$ratio)
data$ratio_ptilerank <- round(pnorm(data$ratio_zscore)*100,2)
```

12) 
```{r}
error <- qt(.975,df=length(data$ratio))*sd(data$ratio)/sqrt(length(data$ratio))
mean(data$ratio)+error
mean(data$ratio)-error
mean(data$ratio)

t.test(data$ratio)
```

14) Run the following code block and answer the following questions. (Type your answers below.)
a) What are the levels of the variable data$nomnial? - 
b) What is the logical result when you test to see if the data$nominal variable is male for the observations that are female? - 
c) What is the logical result when you test to see if the data$nominal variable is male for the observations that are male? - 
d) What is the result when you add the as.numeric() function to the test? - 
```{r}
data$nominal
temp <- data$nominal=="male"
temp
temp <- as.numeric(data$nominal=="male")
temp
```

15) Attach the values in the temp variable to the data frame by running the following code block.

```{r}
data$gender_recode <- temp

class(data$gender_recode)


```

16) Run the following code block to convert the new gender_recode variable to a factor. Examine the data frame using dfSummary() and answer the following question.
a) What is the class of data$gender_recode? - 
b) What are the levels of data$gender_recode? - 
```{r}
data$gender_recode <- factor(data$gender_recode)

view(dfSummary(data))
```

17) Change your working directory by using the menus in RStudio. RStudio - Session - Set Working Directory - Choose Directory... Navigate to the directoy where you want to save the assignments for this class. Save your data frame by running the following code block. After you run the code block, navigate to the folder and make sure the new file has been saved.
```{r}
write.csv(data, file = "Week2data.csv")
```

While you are working on your analysis, your coworker asks you for some help getting some descriptive statistics for a set of numbers and building a boxplot. The data are: [2,4,6,8,10]. Your co-worker read of a useful function for descriptive statistics, but wasn't sure how to get the package to run. The function is called stat.desc() and it's in the pastecs package. Run the following code block to help your coworker. Help your coworker answer the following questions.
a) What is the mean? - 
b) What is the median? - 
c) What is the minimum? - 
d) What is the maximum? - 
e) What is the range? - 
f) What is the first quartile? - 
g) What is the third quartile? - 
```{r}
#install.packages("pastecs") # uncomment this line if you need to install package

library(pastecs) # loads the package into memory

temp1 <- c(2,4,6,8,10) # puts the numbers into a variable
temp1 # shows the values in the variable

stat.desc(temp1) # get descriptive statistics for the variable
summary(temp1) # get summary statistics for the variable.

boxplot(temp1) # creates a box plot for the variable
```


18) Save this assignment as an Rmd file by using the RStudio menus. RStudio - File - Save As... Name the file Week2completedYourName. Turn in your Rmd file and your csv file you just saved.